NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-03-24 main.tex
Update text

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, provides programmatic access to all simulation data in the NMRlipids databank. This enables wide range of novel data-driven applications from construction of machine learning models that predict membrane properties to automatic analyses of virtually any property from all simulations in the databank. Practical implementation of such analyses is illustrated in the flowchart in Fig.~\ref{fig:quality}D. First, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can be then automatically analyzed with the help of README.yaml and mapping files that affiliate specific naming conventions in each simulation to the universal molecule and atom names. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. When {\it Application layer} repositories are organized by mimicking the {\it Databank layer} structure, they can be accessed programmatically and further analyzed using the tools in NMRlipids databank-API by implementing the flowchart demonstrated in Fig.~\ref{fig:quality}E. Novel analyses demonstrating the power of the NMRlipids databank in selecting best simulation models, analyses of rare phenomena and extending MD simulations to new fields are implemented in an {\it application layer} repository located at~\url{www.github.com/NMRLipids/DataBankManuscript}.

@@

\subsection{Applications of the NMRlipids databank}
\subsubsection{Selection of simulation parameters using the NMRlipids databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
The quality of lipid bilayer MD simulations has to be carefully assessed in their applications to minize the detrimental consequences of artificial MD simulation results for lipid bilayers~\cite{antila22b}. This is possible, for example, by using C--H bond order parameters from NMR and X-ray scattering form factors \cite{ollila16}, although it requires comparisons between large number of simulations which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. To streamline this process we have defined quantitative quality measures for lipid bilayer simulations that enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Conformational ensembles of individual lipid molecules are evaluated in the NMRlipids databank by calculating the probabilities for each C-H bond order parameter to locate within experimental error. Furthermore, the quality against X-ray scattering experiments is estimated from the distance of the first form factor minimum between the experiment and simulation. These measures can be used to evaluate membrane dimensions as the form factor minima locations and acyl chain order parameters correlate with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}. 
%while correlating well also with the membrane thickness and lateral packing as shown in Fig.~\ref{fig:quality}F. 
%The X-ray scattering form factors are related to membrane dimensions \textit{via} the electron density profile. 
Here we demonstrate how automatic simulation quality evaluation and NMRlipids databank-API enable rapid selection of the best models for simulations of membranes with POPC and POPE lipids that are the two most biologically abundant neutral membrane lipids~\cite{vanmeer08}.

Predictions for the lateral packing of POPC and POPE membranes in terms of area per lipid, one of the most important parameter used to characterize cellular membranes, diverge between different force fields as illustrated in Fig.~\ref{fig:quality}A. To select the most realistic models among these, we ranked the simulations according to the quality of C-H bond order parameters evaluated against experimental data in Fig.~\ref{fig:top50simulations}. Because the membrane packing correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality}G), force fields occurring in Fig.~\ref{fig:quality}A are ranked according to the $\textit{sn}$-1 chain quality in Figs.~\ref{fig:quality}B and C. Also the highest ranked simulation (POPC bilayer with OPLS3e parameters) is included for the reference.
%for this part of POPE and POPC lipids are considered to give the most realistic results in Fig.~\ref{fig:POPCPOPEapls}C. To this end, the ranking based on C--H bond order parameter quality for the best performing POPC and POPE simulations in the NMRlipids databank are shown in Figs.~\ref{fig:quality}A and B, respectively. 
Direct comparisons with the experimental data are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI} for the most relevant simulations. Based on these rankings, Slipids simulation gives the most realistic prediction for POPC membrane in Fig.~\ref{fig:quality}A with a clear difference to others, and 2\textsuperscript{nd} most realistic prediction for POPE with only marginally lower quality than GROMOS-CKP simulations. Simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayer for POPC (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}), while the area per lipid for POPC is overestimated in Lipid17. For POPE, CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evaluation based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for applications where membrane packing is relevant.

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality5.pdf}
    \caption{ A) Best simulations ranked based on overall order parameter quality.
    B) Best simulations ranked based on the overall order parameter quality of POPE lipid. 
    C)--E) Evaluation against experimental (NMR order parameters and X-ray scattering ) data exemplified for a simulation with the best overall order parameter quality (C), the best quality for POPE lipid (D), and the headgroup quality for POPE (E).
    F) Scatter plots and Pearson correlation coefficients, $r$, for the membrane area per lipid, thickness, first minimum of X-ray scattering form factor and average order parameter of the {\textit sn}-1 acyl chain extracted from the NMRlipids databank. All correlation coefficients have p-value below 0.001 as indicated by $^{***}$. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

@@

Using workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A, we first calculated the cholesterol flip--flop rates from all the simulations available in the NMRlipids databank. Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol). The observed cholesterol flip--flop rates, ranging between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$, are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9$\cdot$10$^{-6}$\,\textmu{}s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Thus, available data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we used the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B to analyse how flip--flop rates calculated from the NMRlipids databank depend on membrane properties. Cholesterol flip--flop rates and their histograms are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B--D. The results reveal a non-linear correlation between cholesterol flip--flop rate and membrane packing which is depicted as area per lipid. Flip--flop rates increase by an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip--flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip--flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The resulting non-zero values range between 0.3 and 322\,\textmu{}m/s with the mean and median of 14\,\textmu{}m/s and 8\,\textmu{}m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%, although weak decrease for the latter two may be visible.
%
%
%
%%Several models on how water dynamics depends on membrane properties have been proposed based on different experimental and %%theoretical methods, yet consensus 
%%%on how water permeability depends on membrane properties, such as area per lipid and thickness, or molecular composition, 
%%has not been reached~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\mathrm{\perp}/D_{||}$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\mathrm{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. 
%Large anisotropy values are understandable as simulations give slightly slower permeation rates and higher lateral diffusion rates than experiments. 
Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\mathrm{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enables the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations.
%can be used to select the best parameters for simulations of large biomolecular complexes and to guide optimizations of new force field parameters for both atomistic and coarse grained simulations. 
Such practises fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amount of MD simulation data via NMRlipids databank-API enables wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip--flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields as demonstrated here by analysing the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}) that is essential parameter in pharmacokinetic modeling and MRI imaging~\cite{nitsche19,topgaard20}. Furthermore, the NMRlipids databank-API delivers access to a training set data that paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict proteins structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientist who do not typically have access to large scale MD simulation data.
%and for analyses of rare phenomena that require exceptional amounts of simulation data.
%On the other hand, universal naming convention enables systematic comparisons between simulations having different model parameters and thereby different naming conventions.
%
%The NMRlipids databank will be of great benefit to scientists involved in MD simulations. They will be able to rapidly evaluate the quality of MD simulations in order to filter out potentially misleading results or facilitate force field parameter development~\cite{antila22b}. The selection of best models for a specific application is demonstrated for PC/PE lipid mixtures in Fig.~\ref{fig:quality}, and previously for PC/PS lipid mixtures~\cite{antila22b}, and lipid headgroups~\cite{bacle21}. 
%
Different types of expected applications of the NMRlipids databank in wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the NMRlipids databank.

@@

\subsection{Adding data into the NMRlipids databank}
The NMRlipids databank is open for additions of simulation data by anyone. The first step is to create an info.yaml file containing the information that needs to be manually entered as listed in Table~\ref{tab:READMEkeys}. This file can be then added into \texttt{/Scripts/BuildDatabank/info\_files} folder in the NMRlipids databank repository via pull requests. After the pull request is manually accepted, the rest of the information for the README.yaml file, listed in Table~\ref{tab:READMEkeys}, will be automatically extracted using the \texttt{/Scripts/BuildDatabank/AddData.py} script. Currently, the NMRlipids databank is composed of simulations found from Zenodo repository with an appropriate licence. Most, but not all, of these trajectories originate from previous NMRlipids projects~\cite{botan15,catte16,antila19,bacle21}.

>>>>

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, provides programmatic access to all simulation data in the NMRlipids databank. This enables wide range of novel data-driven applications from construction of machine learning models that predict membrane properties to automatic analyses of virtually any property from all simulations in the databank. Practical implementation of such analyses is illustrated in the flowchart in Fig.~\ref{fig:quality}D. First, raw data of each simulation can be accessed by iterating through the README.yaml files in the {\it databank layer} that contain links pointing to the locations of simulation trajectories and other relevant files. Simulations can be then automatically analyzed with the help of README.yaml and mapping files that affiliate specific naming conventions in each simulation to the universal molecule and atom names. Finally, the analysis results are stored using the same structure as in the {\it databank layer}. Examples of codes that utilize NMRlipids databank-API and a template for new analyses can be found from locations listed in table~\ref{tab:codes}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. When {\it Application layer} repositories are organized by mimicking the {\it Databank layer} structure, they can be accessed programmatically and further analyzed using the tools in NMRlipids databank-API by implementing the flowchart demonstrated in Fig.~\ref{fig:quality}E. Novel analyses demonstrating the power of the NMRlipids databank in selecting best simulation models, analyses of rare phenomena and extending MD simulations to new fields are implemented in an {\it application layer} repository located at~\url{www.github.com/NMRLipids/DataBankManuscript}. The related codes are listed in table~\ref{tab:codesApplication}.

@@

\subsection{Applications of the NMRlipids databank}
\subsubsection{Selection of simulation parameters using the NMRlipids databank: Best models for most abundant neutral membrane lipids}
%Quality of membrane simulations with different force fields have been evaluated against experimental data during parameterization and in separate comparison studies~\cite{botan15,ollila16,catte16,pluhackova16,perez17,leonard19,??}, but the lack of universal quality measure for membrane simulations complicates the estimations of reliability of simulations. 
%
%For example, OPLS3e parameters overcome CHARMM36 in structural quality for POPC but predicts overestimated ion binding~\cite{kurki22}. On the other hand, CHARMM36 gives the best description for lipid headgroup conformational ensembles~\cite{bacle21}, while GROMOS-CKP parameters best capture the membrane packing in POPS bilayers~\cite{antila22b}. 
%
The quality of lipid bilayer MD simulations has to be carefully assessed in their applications to minize the detrimental consequences of artificial MD simulation results for lipid bilayers~\cite{antila22b}. This is possible, for example, by using C--H bond order parameters from NMR and X-ray scattering form factors \cite{ollila16}, although it requires comparisons between large number of simulations which is laborious even with collaborative approaches~\cite{botan15,catte16,antila19,bacle21}. To streamline this process we have defined quantitative quality measures for lipid bilayer simulations that enable automatic ranking of lipid bilayer simulations based on their quality against experiments. Conformational ensembles of individual lipid molecules are evaluated in the NMRlipids databank by calculating the probabilities for each C-H bond order parameter to locate within experimental error. Furthermore, the quality against X-ray scattering experiments is estimated from the distance of the first form factor minimum between the experiment and simulation. These measures can be used to evaluate membrane dimensions as the form factor minima locations and acyl chain order parameters correlate with the membrane lateral packing and thickness as shown in Figs.~\ref{fig:quality}G,~\ref{fig:QualityCorrelationsSI}, and~\ref{fig:sizedependence}. In addition, the ergodicity of conformational sampling of lipids is estimated by calculating the convergence time of the slowest principal component divided by the length of a simulation. Here we demonstrate how automatic simulation quality evaluation and NMRlipids databank-API enable rapid selection of the best models for simulations of membranes with POPC and POPE lipids that are the two most biologically abundant neutral membrane lipids~\cite{vanmeer08}.

Predictions for the lateral packing of POPC and POPE membranes in terms of area per lipid, one of the most important parameter used to characterize cellular membranes, diverge between different force fields as illustrated in Fig.~\ref{fig:quality}A. To select the most realistic models among these, we first ranked all simulations in the databank according to the quality of C-H bond order parameters evaluated against experimental data in Fig.~\ref{fig:top50simulations}. From these results, we then picked the force fields occurring in Fig.~\ref{fig:quality}A and ranked them according to the quality of $\textit{sn}$-1 chain in Figs.~\ref{fig:quality}B and C. Simulations with $\tau_{\rm rel}$ clearly above one were discarded. Also the highest ranked simulation (POPC bilayer with OPLS3e parameters) is included for the reference. Because the membrane packing correlates with the average order parameter of the $\textit{sn}$-1 chain (Fig.~\ref{fig:quality}G), rankings in Figs.~\ref{fig:quality}B and C were used to select the simulations giving most realistic results in~Fig.~\ref{fig:quality}A. Also direct comparisons with the experimental data for the most relevant simulations are shown in Figs.~\ref{fig:quality}C--E and~\ref{fig:POPC_POPE_dataSI}. Based on these rankings, Lipid17 and Slipids simulations give the most realistic predictions for POPC membrane, while simulations with CHARMM36 and GROMOS-CKP parameters predict overly packed bilayers (overestimated order in Fig.~\ref{fig:POPC_POPE_dataSI}). On the other hand, GROMOS-CKP and Slipids give the most realistic results for POPE, while CHARMM36 and Lipid17 predict membranes that are too packed. In conclusion, the quality evaluation based on the NMRlipids databank suggests that Slipids parameters are the best available choice for simulations with PC and PE lipids, at least for applications where membrane packing is relevant.

\begin{figure}[!t]
    \centering
    \includegraphics[width=\linewidth]{Figures/quality5.pdf}
    \caption{ A) Area per lipids of POPC and POPE lipid bilayers predicted by different force fields at 310\,K in simulations that are available in the NMRlipids databank. The points from the best performing simulations based on rankings in B and C are surrounded by black circles. 
    B) Best POPC simulations ranked based on {\it sn}-1 acyl chain order parameter quality.
    C) Best POPE simulations ranked based on {\it sn}-1 acyl chain order parameter quality. 
    D)--F) Direct comparison against experimental (NMR order parameters and X-ray scattering ) data exemplified for a simulation with the best overall order parameter quality (D), the best quality for POPE lipid (E), and the headgroup quality for POPE (F).
    G) Scatter plots and Pearson correlation coefficients, $r$, for the membrane area per lipid, thickness, first minimum of X-ray scattering form factor and average order parameter of the {\textit sn}-1 acyl chain extracted from the NMRlipids databank. All correlation coefficients have p-value below 0.001 as indicated by~$^{***}$. For more correlations see Fig.~\ref{fig:QualityCorrelationsSI}.
    }
    \label{fig:quality}
\end{figure}

@@

Using workflow depicted in Fig.~\ref{fig:POPCPOPEapls}D, we first calculated the cholesterol flip--flop rates from all the simulations available in the NMRlipids databank. Flip--flops were observed for cholesterol, % (634 events in 77 different simulations), 
DCHOL (18,19-di-nor-cholesterol), %, 16 flip-flops in 1 simulation) 
DOG (1,2-dioleoyl-\textit{sn}-glycerol), %, 44 flip-flops in 4 simulations) 
and SDG (1-stearoyl-2-docosahexaenoyl-\textit{sn}-glycerol). The observed cholesterol flip--flop rates, ranging between 0.001--1.6\,\textmu{}s$^{-1}$ with the mean of 0.16\,\textmu{}s$^{-1}$ and median of 0.07\,\textmu{}s$^{-1}$, are in line with the previously reported values from atomistic MD simulations~\cite{gu19,javanainen19,baral20}. The flip--flop rate of DCHOL, 0.2\,\textmu{}s$^{-1}$, was close to the average value of cholesterol, while average rates for diacylglycerols DOG and SDG were higher than for cholesterol, 0.4\,\textmu{}s$^{-1}$ and 0.5\,\textmu{}s$^{-1}$, respectively. Flip--flops were not observed for other lipids, giving the upper limits for PC lipid flip--flop rate of 9$\cdot$10$^{-6}$\,\textmu{}s$^{-1}$ and for ceramide (\textit{N}-palmitoyl-\textsc{D}-\textit{erythro}-sphingosine) of 0.002\,\textmu{}s$^{-1}$. Thus, available data in the NMRlipids databank suggests that the lipid flip--flip rate decreases in the order of diacylglycerols $>$ cholesterol $>$ other lipids including ceramides. However, the amount of data for diacylgycerols (8 simulations with Lipid17 force field) and ceramide (3 simulations with CHARMM36 force field) is less than for cholesterol (83 simulations), thus we cannot fully exclude the effect of force field or composition on this comparison.

Nevertheless, we used the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}E to analyse how flip--flop rates calculated from the NMRlipids databank depend on membrane properties. Cholesterol flip--flop rates and their histograms are plotted as a function of membrane thickness, lateral density and order in Figs.~\ref{fig:flip-flops} B--D. The results reveal a non-linear correlation between cholesterol flip--flop rate and membrane packing which is depicted as area per lipid. Flip--flop rates increase by an order of magnitude when membrane packing density decreases and a major jump is observed at low membrane packing. The order of magnitude changes in cholesterol flip--flop rate with the membrane composition may have major implications in understanding lipid trafficking and membrane biochemistry~\cite{gu19,baral20}. Because the results from the NMRlipids databank are averaged over large range of membrane compositions and force fields, they show that the strong dependence of cholesterol flip--flop rate on membrane properties is not limited to certain lipid compositions or force fields used in previous studies~\cite{gu19,javanainen19,baral20}.

@@

To this end, we first calculated the water permeability through membranes from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The resulting non-zero values range between 0.3 and 322\,\textmu{}m/s with the mean and median of 14\,\textmu{}m/s and 8\,\textmu{}m/s, respectively. These values agree with the previously reported simulation results~\cite{venable19,camilo2022}, but are on average slightly larger than experimental values reported for PC lipids in the liquid crystalline phase,~0.19--0.33\,\textmu{}m/s~\cite{jansen95}. Using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B, we then plotted the observed permeabilities and their histogrammed values in Figs.~\ref{fig:permeability}B--E as a function of temperature, membrane thickness, area per lipid, and acyl chain order. As expected, the permeability increases with the temperature, giving an average energy barrier of 17\,$\pm$\,4\,$k_\mathrm{B}T$ for the water permeation from the Arrhenius plot in Fig.~\ref{fig:permeability}B. On the other hand, the permeability of water decreases on average when membranes become more packed, \textit{i.e.}, with decreasing area per lipid and increasing thickness and acyl chain order (Figs.~\ref{fig:permeability} C--E). Permeation of water through bilayers depends on membrane properties also according to previous studies, but there is no established consensus on whether the area per lipid~\cite{nagle08} or bilayer thickness~\cite{frallicciardi22} is the main parameter determining the permeability. 
%is previously reported from simulations and experiments for some systems, but not for all~\cite{mathai07,venable19,frallicciardi22}. 
Our analysis over the NMRlipids databank, containing significantly more data than what was available in previous studies, suggest non-linear dependencies on both of these parameters, yet the linear correlation would be a good approximation for thicknesses above $\sim$\,3.9\,nm and area per lipids below $\sim$69\,\AA{}$^2$ (insets in Figs.~\ref{fig:permeability} C--D).
%within the range of experimentally reported values  permeability which is 1-2 orders of magnitude slower than osmotic permeability \cite{jansen95,??}. 
Clear dependencies of permeability on charged lipids, cholesterol, POPE, or hydration level were not observed in Fig.~\ref{fig:permeationSI} in the supplementary information.
%
%%Several models on how water dynamics depends on membrane properties have been proposed~\cite{mathai07,nitsche13,nitsche16,shinoda16,venable19,frallicciardi22}.

To analyze how water diffusion anisotropy depends on membrane properties in a multi-lamellar lipid bilayer system, we calculated the water diffusion parallel to the membrane surface from all simulations in the NMRlipids databank using the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}A. The parallel diffusion coefficient of water, $D_{||}$, decreases with reduced hydration and increases with the temperature, but dependencies on membrane area per lipid, thickness, or fraction of charged lipids were not observed in Figs.~\ref{fig:permeability} and~\ref{fig:diffusionSI} when the workflow depicted in Fig.~\ref{fig:POPCPOPEapls}B was used. Simulation results are close to the experimental values with low hydration levels in Fig.~\ref{fig:permeability} F, but increase to approximately two times higher than the experimental value for bulk water diffusion value (3.1$\cdot$\,10$^{-9}$\,m$^2$/s at 313\,K~\cite{khakimov08}) with high hydration levels. This is not surprising as the most common water model used in membrane simulations, TIP3P, overestimates the bulk water diffusion~\cite{pathirannahalage21}. To estimate the diffusion anisotropy of water, $D_\mathrm{\perp}/D_{||}$, in multilamellar membrane system, the permeability coefficients of water through membranes were translated to perpendicular diffusion coefficients, $D_\mathrm{\perp}$, using the Tanner equation~\cite{tanner78,wasterby02}. The resulting perpendicular diffusion coefficients are approximately five orders of magnitude smaller than lateral diffusion coefficients of water (Figs.~\ref{fig:permeability} G and H), which is at the upper limit of the anisotropy estimated from the experimental data~\cite{nitsche19}. Significant increase in the diffusion anisotropy with membrane packing is observed, as $D_\mathrm{\perp}/D_{||}$ drifts away from one with decreasing area per lipid and increasing thickness in Figs.~\ref{fig:permeability} G and H. This follows from decreasing water permeability with membrane packing (Figs.~\ref{fig:permeability} C and D), while lateral diffusion remains approximately constant (Figs.~\ref{fig:diffusionSI} A and C).

@@

The focus of biomolecular simulations is moving from studies of individual molecules to larger complexes and even whole cells and organelles~\cite{johnson15,thornburg22,gupta22}. Simultaneously, machine learning based models predicting behaviour of biomolecules and automatic approaches to parametrize models are emerging~\cite{jumper21,antila22b}. The resources delivered by the NMRlipids databank will support the development in both of these directions. Automatic quality evaluation and ranking of simulations against experimental data enables the selection of best simulations for specific applications without laborious manual force field evaluation. This also streamlines automatic parametrization procedures for atomistic and coarse grained simulations. Such practises fostering the accuracy of simulations are becoming increasingly important due the accumulation of small errors when complexity and size of simulated systems are increasing.

The open programmatic access to unprecedented amount of MD simulation data via NMRlipids databank-API enables wide range of novel data-driven analyses and machine learning applications. The analyses of cholesterol flip--flop events (Fig.~\ref{fig:flip-flops}) and water permeation through membranes (Fig.~\ref{fig:permeability}) demonstrate how large amount of accessible simulation data in terms of quantity (\textit{e.g.}, simulation length and number of conformations) and content (\textit{e.g.}, lipid compositions and ion concentrations) enable analyses of rare phenomena that are beyond the current possibilities for a single research group. Such analyses also pave the way for applications of MD simulations in new fields as demonstrated here by analysing the anisotropic diffusion of water in membrane systems (Fig.~\ref{fig:permeability}) that is essential parameter in pharmacokinetic modeling and MRI imaging~\cite{nitsche19,topgaard20}. Furthermore, the NMRlipids databank-API delivers access to a training set data that paves the way for diverse machine learning applications to 
%For example, a machine learning model predicting electron density profiles from form factors would be highly useful in interpretation of scattering experiments.
%In addition, the data contained in the NMRlipids databank will enable training of more elaborated models to
predict membrane properties. Such applications are analogous to AlphaFold \cite{jumper21} and other tools that predict proteins structures from their sequence using artificial intelligence. These possibilities are particularly valuable for scientist who do not typically have access to large scale MD simulation data. Different types of expected applications of the NMRlipids databank in wide range of fields are listed in Table~\ref{tab:applications}, yet the scope of such applications
%in biology, biotechnology, material science, biomolecular imaging and beyond
is expected to further widen with increasing amount of data in the NMRlipids databank.

@@

\subsection{Adding data into the NMRlipids databank}
The NMRlipids databank is open for additions of simulation data by anyone. The list information that a contributor has to deliver is given in Table~\ref{tab:READMEkeys}. The rest of the information to be stored in README.yaml files, also listed in Table~\ref{tab:READMEkeys}, will be automatically extracted using the \texttt{/Scripts/BuildDatabank/AddData.py} script. In practise, the manually entered data is first stored into an info.yaml file that is then added into \texttt{/Scripts/BuildDatabank/info\_files} folder via pull request. To avoid ineligible entries and minimize human errors, the pull requests are monitored before the acceptance and generation of README.yaml files.  Currently, the NMRlipids databank is composed of simulations that are found from the Zenodo repository with an appropriate licence. Most, but not all, of these trajectories originate from previous NMRlipids projects~\cite{botan15,catte16,antila19,bacle21}.

\subsection{Experimental data}
Experimental data used in the quality evaluation, currently composed of C--H bond order parameters and X-ray scattering form factors, are stored in \texttt{/Data/experiments} in the NMRlipids databank repository. Similarly to simulations, each experimental data set has a README.yaml file containing all the relevant information about the experiment. The keys and their descriptions for the experimental data are given in Table~\ref{tab:READMEkeysEXP}. NMR data currently in the NMRlipids databank are taken from Refs.~\citenum{scherer87,ferreira13,antila19,melcr20,bacle21} and X-ray scattering data from Refs.~\citenum{Kucerka08a,kucerka11,pan12b,pan14,kucerka15,NMRlipidsIII}. In addition, previously unpublished NMR data for POPE, POPG and DOPC was acquired as described in the supplementary information (Figs.~\ref{POPEexp1}-\ref{fig:DOPCexp}) and contributed to the databank.

\subsection{Analysing simulations}
In practise, simulation data in the NMRlipids databank can be currently analyzed by first downloading the data using the information in README.yaml and then performing the analysis on local computer utilising the universal naming conventions for molecules and atoms therein. This procedure is illustrated in Fig.~\ref{fig:overlay}D and tables~\ref{tab:codes} and~\ref{tab:codesApplication} list examples of codes that perform such analyses in practise.
